---
title: "Tidy"
editor: 
  markdown: 
    wrap: sentence
---

This section performs the complete data cleaning and standardization of the Olympic results, career progression data, IAAF 2005 scoring table (iaaf-2005.json), and the world-class standards table.
The objective is to produce a unified, well-structured dataset containing each athlete’s best annual performance, age at performance, event metadata, and standardized IAAF points.

Four key tidy tables are produced for use in downstream analysis:

-   **Olympics_data:** cleaned athlete, event, and Games metadata

-   **Career_progression:** cleaned and date-formatted yearly performance records

-   **iaaf_table:** flattened IAAF reference table in tidy format

-   **wac_standards:** standardized world-class standards in long format

## Clean `Olympics_data.csv`

The Olympic results file is processed to create a clean athlete-level metadata table, ensuring consistent formats, corrected anomalies, and standardized event classifications.
This cleaned dataset will later be joined with career progression records to enrich performance analysis.

**Key transformations:**

-   Read the raw CSV of Olympic track & field results.

-   `clean_names()` to lowercase and snake_case all column names.

-   Split date spans (e.g., `"01 Aug 2012 – 12 Aug 2012"`) into `start` and `end` dates.

-   Split event strings into gender (`"Men"`/`"Women"`) and event name, keeping remaining tokens intact.

-   Remove wheelchair events.

-   Correct one known birthdate anomaly (`Aminata CAMARA`).

-   Normalize gender labels to `"Men"`/`"Women"`.

-   Convert `birth_date` and `end` to `Date` objects (`%Y-%m-%d`).

-   Calculate `age` at Games end date.

-   Recode `games` as an ordered factor with readable labels (e.g., `"Tokyo '20"`).

-   Classify each event into `event_type` categories (Sprints, Middle Distance, Throws, etc.).

-   Assign `event_category` as `"Track"` or `"Field"`.

-   Normalize legacy event labels (e.g., `"Javelin Throw (old)" → "Javelin Throw"`).

-   Extract `games_year` from the Games end date.

-   Rename `nat` to `nationality`.

This process produces a clean metadata table, `Olympics_data_tidy`

```{r, echo = TRUE}
Olympics_data <- read_csv("/Users/yushiliu/Stat 468 final project/data/track and field data/olympics_data.csv", show_col_types = F) %>%
  janitor::clean_names() %>% 
  separate_wider_delim(date, delim = "–", names = c("start", "end")) %>%
  separate_wider_delim(cols = event, names = c("gender", "event"), delim = " ",
                           too_many = "merge") %>%
  filter(str_detect(event, "Wheelchair", negate = T)) %>% 
  mutate(
        # Manually fixed Aminata CAMARA's age
    birth_date = if_else(name == "Aminata CAMARA", "06 DEC 1973", birth_date),
    gender = if_else(gender == "Men's", "Men", "Women"),
    across(c(birth_date, end),  ~as.Date(format(
    as.Date(., format = "%d %b %Y"), "%Y-%m-%d"))),
    age = year(end) - year(birth_date),
         games = factor(games, ordered = T,
                        levels = c("The XXXII Olympic Games",
                                   "The XXXI Olympic Games",
                                   "The XXX Olympic Games",
                                   "The XXIX Olympic Games",
                                   "The XXVIII Olympic Games",
                                   "The XXVII Olympic Games",
                                   "The XXVI Olympic Games"
                        ),
                        labels = c("Tokyo '20", "Rio '16", "London '12", 
                                   "Beijing '08", "Athens '04", "Sydney '00",
                                   "Atlanta '96")),
    event_type = case_match(
      event,
      c("100 Metres", "200 Metres", "400 Metres", "400 Metres Hurdles", "100 Metres Hurdles", "110 Metres Hurdles") ~ "Sprints",
      c("800 Metres", "1500 Metres", "3000 Metres Steeplechase") ~ "Middle Distance",
      c("5000 Metres", "10,000 Metres") ~ "Long Distance",
      c("Heptathlon", "Decathlon") ~ "Combined Events",
      c("High Jump", "Long Jump", "Triple Jump", "Pole Vault") ~ "Jumps",
      c("Shot Put", "Discus Throw", "Hammer Throw", "Javelin Throw", "Javelin Throw (old)") ~ "Throws",
      c("10 Kilometres Race Walk", "20 Kilometres Race Walk", "50 Kilometres Race Walk", "Marathon") ~ "Road Races",
      .default = "Other"
    ),
         event_category = if_else(str_detect(event, "Metres|Walk|Wheelchair") |event %in% c("Marathon"), "Track", "Field")
         ) %>% 
  mutate(event = if_else(event == "Javelin Throw (old)", "Javelin Throw", event),
         games_year = year(end)) %>% rename("nationality" = nat)
```

## Clean `Career_Progression.csv`

The career progression dataset is cleaned to provide yearly performance records.\

Key steps:

-   Removing duplicate rows.

-   Parsing competition dates.

-   Flagging retirement status (`retired = TRUE` if last competition year ≤ 2022).

-   Calculating **training age** (number of years since first recorded season) for each athlete-event.

The result is stored as `Career_progression`.

```{r}
# load career progression data and identify athletes who are still active
Career_progression <- read_csv("../data/track and field data/career_progression (2).csv", show_col_types = F) %>% 
  distinct() |> 
  mutate(date = parse_date_time(date, orders = c("%d %b %Y", "%d-%b-%y"))) %>%
  mutate(retired = if_else(max(year) <= 2022, T, F),
         training_age = year - min(year) + 1,
         .by = c("athlete_link", "event")) 
```

## Flatten the IAAF Reference Table (`iaaf-2005.json`)

1.  Reading the JSON file containing IAAF scoring data.

2.  Flattening the nested structure into a tidy table with columns: `gender`, `event`, `mark`, and `points`.

    This creates the reference table `iaaf_table`.

```{r, echo=TRUE}
raw_json <- read_json("../data/track and field data/iaaf-2025.json", simplifyVector = FALSE)

# Convert list of lists to a flat tibble
iaaf_table <- tibble::tibble(
  gender = sapply(raw_json, `[[`, "gender"),
  event = sapply(raw_json, `[[`, "event"),
  mark = sapply(raw_json, `[[`, "mark"),
  points = sapply(raw_json, `[[`, "points")
)
```

## Tidy world-class standards(`wac-standards`)

-   Reshaping the data into long format with `gender` as a variable.

-   Standardizing event names to match IAAF vocabulary.

-   Converting all time-based marks to seconds (HH:MM:SS → seconds).

-   Adjusting women’s hurdles labels to `100mH`.

The result is stored as `wac_standards`.

```{r}
#compute world class standard and take mean of men and women
wac_standards <- tribble(
  ~Men, ~Event, ~Women,
  "10.00", "100m", "11.07",
  "20.16", "200m", "22.57",
  "44.85", "400m", "50.75",
  "1:44.50", "800m", "1:59.00",
  "3:33.00", "1500m", "4:01.50",
  "13:01.00", "5000m / 5km road", "14:50.00",
  "27:00.00", "10,000m / 10km road", "30:20.00",
  "2:06:30", "Marathon", "2:23:30",
  "8:15.00", "3000m SC", "9:18.00",
  "13.27", "110mH / 100mH", "12.73",
  "48.50", "400m H", "54.65",
  "2.33", "High Jump", "1.97",
  "5.82", "Pole Vault", "4.73",
  "8.27", "Long Jump", "6.86",
  "17.22", "Triple Jump", "14.55",
  "21.50", "Shot Put", "18.80",
  "67.50", "Discus Throw", "64.50",
  "78.20", "Hammer Throw", "74.00",
  "85.50", "Javelin Throw", "64.00",
  NA, "Heptathlon", "6500", #points 1174
  "8550", "Decathlon", NA, #1212
  "1:19:20", "20km Race Walk", "1:29:00",
  "2:28:00", "35km Race Walk", "2:48:00")

wac_standards <- wac_standards |> 
  mutate(Event = case_match(
    Event,
    "5000m / 5km road"       ~ "5000m",
    "10,000m / 10km road"    ~ "10000m",
    "110mH / 100mH"          ~ "110mH",  
    "400m H"                 ~ "400mH",
    "High Jump"              ~ "HJ",
    "Pole Vault"             ~ "PV",
    "Long Jump"              ~ "LJ",
    "Triple Jump"            ~ "TJ",
    "Shot Put"               ~ "SP",
    "Discus Throw"           ~ "DT",
    "Hammer Throw"           ~ "HT",
    "Javelin Throw"          ~ "JT",
    "20km Race Walk"         ~ "20,000mW",
    "35km Race Walk"         ~ "35,000mW",
    "Marathon"               ~ "Road Marathon",
    "Heptathlon"             ~ "Hept.",
    "Decathlon"              ~ "Dec.",
    .default = Event
  ))

#convert time to seconds
parse_mark_to_seconds <- function(mark_vec) {
  sapply(mark_vec, function(mark) {
    if (is.na(mark)) return(NA_real_)
    parts <- unlist(strsplit(mark, ":"))
    parts <- as.numeric(parts)
    if (length(parts) == 3) {
      # HH:MM:SS
      return(parts[1]*3600 + parts[2]*60 + parts[3])
    } else if (length(parts) == 2) {
      # MM:SS
      return(parts[1]*60 + parts[2])
    } else {
      # Assume plain seconds
      return(as.numeric(mark))
    }
  })
}

wac_long <- wac_standards |> 
  pivot_longer(
    cols = c(Men, Women),
    names_to = "gender",
    values_to = "mark"
  ) |> 
  mutate(mark = parse_mark_to_seconds(mark), 
         gender = tolower(gender), 
         Event = case_when(
           Event == "110mH" & gender == "women" ~ "100mH",
          TRUE ~ Event))
```

## Summary and Importance

This unified tidy dataset forms the foundation for all subsequent modeling.
By integrating performance records, event metadata, and standardized scoring into a consistent structure, it supports accurate estimation of aging curves through Functional Principal Component Analysis (FPCA) and enables effective clustering of performance trajectories.
The standardized format also ensures reproducibility and makes predictions directly interpretable across athletes, events, and time periods.
