---
title: "Transform"
---

After cleaning and standardizing the datasets, the next step was to derive analytical variables required for modeling. This stage enriches the tidy data by linking metadata, creating features, and mapping performances to standardized scoring systems.

**Key tasks:**

-   Link each performance record to Olympic athlete/event metadata

-   Compute athlete age metrics (age at performance in years and days) and training age

-   Classify events by type and category.

-   Convert performances to numeric marks (seconds for track; numeric values for field).

-   Map marks to IAAF points using a rolling/closest join.

-   Convert world-athlete championship (WAC) to points and compute the mean reference by gender.

## Link career progression to Olympic metadata

First we join the career progression dataset with Olympic athlete metadata (only keeping records with a known birth date), ensures unique athlete-event combinations, and keeps key identifying variables.

It then calculates:

-   age_years: athlete’s age in years at the time of performance

-   age_days: exact age in days

It also removes the "h" suffix from performance to standardize hand-timed results.

```{r, echo=TRUE}
athlete_bests <-
  Career_progression %>%  inner_join(
    Olympics_data %>%  filter(!is.na(birth_date)) %>%
      distinct(birth_date, athlete_links, event, event_type, .keep_all = T) %>%
      select(name, birth_date, athlete_links, event, nationality, event_type, 
             event_category, gender, games_year),
    by = c("athlete_link" = "athlete_links", "event")
  ) %>%  
  mutate(age_years = year(date) - year(birth_date), 
         age_days = as.double(
           difftime(
             date, birth_date,units = "days"
           )
         ),
         performance = str_remove_all(performance, "h")
```

## Classify events, compute performance_age, and convert performance to numeric marks

Events are grouped into broad disciplines (e.g., Sprints, Middle Distance, Jumps) and assigned a Track or Field classification. The athlete’s age at the time of performance is calculated, and the raw performance string is converted into a numeric mark suitable for modeling.

Specifically, the code:

-   Assigns `event_type` (Sprints, Middle Distance, Long Distance, Combined Events, Jumps, Throws, Road Races, or Other) and sets `event_category` to either Track or Field.

-   Computes `age_years` and `age_days`, and removes the "`h`" hand-timed suffix from performance strings.

-   Generates `mark`: for Track events, parses the time string and converts it to seconds using `lubridate::parse_date_time2` and `difftime`; for Field events, extracts a numeric distance using `parse_number`.

```{r, echo=TRUE}
# Classify events
event_type = case_match(
      event,
      c("100 Metres", "200 Metres", "400 Metres", "400 Metres Hurdles", "100 Metres Hurdles", "110 Metres Hurdles") ~ "Sprints",
      c("800 Metres", "1500 Metres", "3000 Metres Steeplechase") ~ "Middle Distance",
      c("5000 Metres", "10,000 Metres") ~ "Long Distance",
      c("Heptathlon", "Decathlon") ~ "Combined Events",
      c("High Jump", "Long Jump", "Triple Jump", "Pole Vault") ~ "Jumps",
      c("Shot Put", "Discus Throw", "Hammer Throw", "Javelin Throw", "Javelin Throw (old)") ~ "Throws",
      c("10 Kilometres Race Walk", "20 Kilometres Race Walk", "50 Kilometres Race Walk", "Marathon") ~ "Road Races",
      .default = "Other"
    ),
         event_category = if_else(str_detect(event, "Metres|Walk|Wheelchair") |event %in% c("Marathon"), "Track", "Field")
mutate(age_years = year(date) - year(birth_date), 
         age_days = as.double(
           difftime(
             date, birth_date,units = "days"
           )
         ),
         performance = str_remove_all(performance, "h") # I will count hand timed results as legitimate
         ) %>%
    mutate(mark = if_else(
      event_category == "Track",
      as.numeric(
        difftime(
          lubridate::parse_date_time2(performance, orders = c("%H:%M:%S", "%M:%S:00", "%M:%OS", "%OS"), exact = T),
          lubridate::parse_date_time2("0", orders ="S"),
          units = "secs"
        )
      ), parse_number(performance)))
```

## Normalize event labels for IAAF scoring

This step standardizes event names and gender labels so they match the IAAF scoring reference, ensuring reliable joins later. It converts gender to lowercase and maps verbose event strings (for example, “100 Metres,” “High Jump,” “10 Kilometres Race Walk”) to the short IAAF codes (for example, 100m, HJ, 10,000mW). Events without a defined mapping are left unchanged. This normalization guarantees consistent keys for scoring and downstream analyses.

```{r,echo=TRUE}
athlete_bests_clean <- athlete_bests |> 
  mutate(
    gender = tolower(gender),
    event = case_when(
      event == "100 Metres" ~ "100m",
      event == "200 Metres" ~ "200m",
      event == "400 Metres" ~ "400m",
      event == "800 Metres" ~ "800m",
      event == "1500 Metres" ~ "1500m",
      event == "5000 Metres" ~ "5000m",
      event == "10,000 Metres" ~ "10000m",
      event == "Marathon" ~ "Road Marathon",
      event == "Triple Jump" ~ "TJ",
      event == "Long Jump" ~ "LJ",
      event == "High Jump" ~ "HJ",
      event == "Shot Put" ~ "SP",
      event == "Discus Throw" ~ "DT",
      event == "Hammer Throw" ~ "HT",
      event == "Javelin Throw" ~ "JT",
      event == "Decathlon" ~ "Dec.",
      event == "Heptathlon" ~ "Hept.",
      event == "Pole Vault" ~ "PV",
      event == "10 Kilometres Race Walk" ~ "10,000mW",
      event == "50 Kilometres Race Walk" ~ "50,000mW",
      event == "20 Kilometres Race Walk" ~ "20,000mW",
      event == "100 Metres Hurdles" ~ "100mH",
      event == "110 Metres Hurdles" ~ "110mH",
      event == "400 Metres Hurdles" ~ "400mH",
      event == "3000 Metres Steeplechase" ~ "3000m SC",
      TRUE ~ event
    )
  )
```

## Map athlete performance marks to IAAF points and join `athlete_bests_clean` with `iaaf-table` to get full dataset

This step assigns IAAF point scores to each performance by joining `athlete_bests_clean` to the IAAF reference table on gender and event, using a rolling “closest mark ≤ observed mark” rule. If no matching score is found, the point value is set to zero. The result, `athlete_bests_points`, is the modeling-ready table with season-best records expressed on a common IAAF points scale.

```{r,echo=TRUE}
# join iaaf_table with athlete_bests_clean and get full dataset with result points
round_decimals <- function(x, digits = 2) {
  round(x * 10^digits) / 10^digits}


athlete_bests_points <- 
  athlete_bests_clean |>
    left_join(iaaf_table, join_by(gender, event, closest(mark <= mark))) |> 
  replace_na(list(points = 0))
```

## Convert WAC standards to points and compute mean references

This step expresses the world-class standards on the same IAAF points scale used for athlete results. It standardizes WAC event labels and times, joins to the IAAF table using a “closest mark ≤ observed mark” rule, and fills missing scores with zero. For combined events, fixed anchors are applied (Heptathlon women = 1174, Decathlon men = 1212). Finally, it computes the mean WAC points by gender to provide simple reference lines for later visualizations.

```{r,echo=TRUE}
wac_points <- wac_standards_tidy %>%
  left_join(
    iaaf_table_tidy,
    dplyr::join_by(gender, Event == event, closest(mark <= mark))
  ) %>%
  tidyr::replace_na(list(points = 0)) %>%
  mutate(points = dplyr::case_when(
    Event == "Hept." & gender == "women" ~ 1174,
    Event == "Dec."  & gender == "men"   ~ 1212,
    TRUE ~ points
  ))

mean_points_by_gender <- wac_points %>%
  group_by(gender) %>%
  summarise(mean_points = mean(points, na.rm = TRUE), .groups = "drop") %>%
  tibble::deframe()
```

## Model-preparation

This step prepares the FPCA input at the athlete–event level. It filters to records with non-missing IAAF points and age, builds a unique identifier by concatenating `athlete_link` and `event`, and selects the three fields required by face.sparse: athlete_id, performance_age (age in years at performance), and iaaf_score (IAAF points). It also creates helper vectors (id and uid) for indexing subjects during modeling.

```{r,echo=TRUE}
# predict on full dataset(population)
# Filter out missing points athletes and rename for clarity
fd_data <- athlete_bests_points |> 
  filter(!is.na(points), !is.na(age_years)) |> 
  mutate(id = paste(athlete_link, event)) |> 
  select(athlete_id = id,
         performance_age = age_years,
         iaaf_score = points)

id <- fd_data$athlete_id
uid <- unique(id)
```

## Key outputs for modeling and Visualization:

-   `athlete_bests_points`: season-best results with IAAF points

-   `wac_points` and `mean_points_by_gender`: world-class reference levels for comparison in plots
