---
title: "Visualize"
---

This section builds visuals on top of the prepared dataset and the sparse FPCA (FACE) fit:

-   Prepare modeling-ready columns (`fd_data`).
-   Fit a sparse FPCA model and visualize a few subject-level aging curves with 95% CIs and WAC reference lines.
-   Show FACE components (eigenfunctions) and a legend for line styles.
-   Plot a single hand-picked athlete.
-   Cluster population-level **percent-change** trajectories, then repeat clustering **by event type**.

Assumes earlier sections created: `athlete_bests_points`, `mean_points_by_gender`.

## Prepare data and fit FACE

```{r}
# predict on full dataset(population)
# Filter out missing points athletes and rename for clarity
fd_data <- athlete_bests_points |> 
  filter(!is.na(points), !is.na(age_years)) |> 
  mutate(id = paste(athlete_link, event)) |> 
  select(athlete_id = id,
         performance_age = age_years,
         iaaf_score = points)

id <- fd_data$athlete_id
uid <- unique(id)
face_input <- data.frame(
  y = fd_data$iaaf_score,
  argvals = fd_data$performance_age,
  subj = id)
#Fit sparse FPCA
fit_face <- face.sparse(face_input, argvals.new = seq(11, 58, by = 1))
#create a data double
data.h <- face_input
tnew <- fit_face$argvals.new
#prepare data for prediction
seq <- seq(11, 58, by = 1)
k <- length(seq)
Sample <- c(5, 8, 12, 20)  # Customize based on your data


for(i in 1:4){
  print(uid[Sample[i]])
  sel <- which(id == uid[Sample[i]])
  dati <- fd_data[sel, ] |> 
  mutate(y = iaaf_score,
       argvals = performance_age,
        subj = athlete_id) |> 
  select(y,argvals, subj)

if (nrow(dati) == 0) {
  message("Skipping subject ", uid[Sample[i]], " — no data.")
  next}

# Create the prediction dataframe
dati_pred <- data.frame(
  y       = rep(NA, nrow(dati) + k),
  argvals = c(rep(NA, nrow(dati)), seq),
  subj    = rep(dati$subj[1], nrow(dati) + k)
)


dati_pred[1:nrow(dati), ] <- dati

yhat <- predict(fit_face, dati_pred)


Ord  <- (nrow(dati) + 1):(nrow(dati) + k)
```

## Sample athlete's aging curves(predicted + CI + population mean + WAC refs)

```{r}
Sample <- which(uid == "https://worldathletics.org/athletes/norway/svein-inge-valvik-014215542 DT")

fit_face <- face.sparse(face_input, argvals.new = seq(11, 58, by = 1))
#create a data double
data.h <- face_input
tnew <- fit_face$argvals.new
#prepare data for prediction
seq <- seq(11, 58, by = 1)
k <- length(seq)
for(i in Sample){
  sel <- which(id == uid[i])
  #dati <- fd_data_top10[sel,] 
  

dati <- fd_data_top10[sel, ] |> 
  dplyr::transmute(y = iaaf_score,
         argvals = performance_age,
         subj = athlete_id)

# Create the prediction dataframe
dati_pred <- data.frame(
  y       = rep(NA, nrow(dati) + k),
  argvals = c(rep(NA, nrow(dati)), seq),
  subj    = rep(dati$subj[1], nrow(dati) + k)
)

dati_pred[1:nrow(dati), ] <- dati

yhat <- predict(fit_face, dati_pred)


Ord  <- (nrow(dati) + 1):(nrow(dati) + k)

#Plot
plot(dati$argvals, dati$y,
     xlab = "Performance Age (years)",
     ylab = "IAAF score",
     pch = 19, col = "blue",
     xlim = range(seq),
     ylim = range(c(dati$y, yhat$y.pred[Ord]), na.rm = TRUE),
     bty = "n")

lines(seq, yhat$y.pred[Ord], col = "red",  lwd = 2)
lines(seq, yhat$y.pred[Ord] + 1.96 * yhat$se.pred[Ord], col = "red", lty = 2)
lines(seq, yhat$y.pred[Ord] - 1.96 * yhat$se.pred[Ord], col = "red", lty = 2)
lines(seq, fit_face$mu.new, col = "black", lwd = 2)

}

#extract mean, covariance, variance, correlation function, and eigenfunctions
#Smooth marginal mean
m <- fit_face$mu.new
#Smooth covariance
Cov <- fit_face$Chat.new
#Pointwise covariance
Cov_diag <- diag(Cov)
#Smooth correlation 
Cor <- fit_face$Cor.new
#Pointwise prediction intervals
m_p_2sd <- m + 2 * sqrt(Cov_diag)
m_m_2sd <- m - 2 * sqrt(Cov_diag)
#Smooth eigenfunctions 
eigenf <- fit_face$eigenfunctions
#Smooth eigenvalues 
eigenv <- fit_face$eigenvalues

#first 3 eigenfunctions
par(mar = c(4.5, 4.5, 0, 2))
col_me <- c("cornflowerblue", "aquamarine3", "coral1")

plot(1, type = "n", xlab = "Performance Age (years)", ylab = "",
     xlim = range(tnew), ylim = range(eigenf), bty = "n")

for(i in 1:3){
  lines(tnew, eigenf[,i], col = col_me[i], lwd = 3)
}

legend("topright", c("PC1", "PC2", "PC3"), 
       col = col_me, lty = 1, bty = "n", lwd = 3)
```

## FACE components: eigenfunctions(PC1-PC3) and line legend

```{r}
#extract mean, covariance, variance, correlation function, and eigenfunctions
#Smooth marginal mean
m <- fit_face$mu.new
#Smooth covariance
Cov <- fit_face$Chat.new
#Pointwise covariance
Cov_diag <- diag(Cov)
#Smooth correlation 
Cor <- fit_face$Cor.new
#Pointwise prediction intervals
m_p_2sd <- m + 2 * sqrt(Cov_diag)
m_m_2sd <- m - 2 * sqrt(Cov_diag)
#Smooth eigenfunctions 
eigenf <- fit_face$eigenfunctions
#Smooth eigenvalues 
eigenv <- fit_face$eigenvalues

#first 3 eigenfunctions
par(mar = c(4.5, 4.5, 0, 2))
col_me <- c("cornflowerblue", "aquamarine3", "coral1")
#plot PC's
plot(1, type = "n", xlab = "Performance Age (years)", ylab = "",
     xlim = range(tnew), ylim = range(eigenf), bty = "n")

for(i in 1:3){
  lines(tnew, eigenf[,i], col = col_me[i], lwd = 3)
}

legend("topright", c("PC1", "PC2", "PC3"), 
       col = col_me, lty = 1, bty = "n", lwd = 3)

par(mfrow = c(1, 1))
plot(1, type = "n", axes = FALSE, xlab = "", ylab = "")  # empty plot
legend("center", 
       legend = c("Predicted Aging Curve", "95% CI", "Population Mean Aging Curve", 
                  "World Class Standard Mean (Men)", "World Class Standard Mean (Women)"),
       col = c("red", "red", "black", "blue", "red"),
       lty = c(1, 2, 1, 3, 3),
       lwd = 2, bty = "n")
```

## Cluster population aging curves based on IAAF-points' percent change

```{r}
#cluster the whole population based on percent change
# set up
id <- fd_data$athlete_id
uid <- unique(id)
#set the grid where to predict
seq <- seq(11, 58, by = 1)
k <- length(seq)
n <- length(uid)
# Set the predicted percent change in IAAF points matrix: one fewer column since we lose first value
PercentChange_mat <- matrix(NA, nrow = n, ncol = k - 1)

# loop over each athlete, predict every curve
for(i in 1:n){
  sel <- which(id == uid[i])
  dati <- fd_data[sel, ] |>
    mutate(y = iaaf_score,
              argvals = performance_age,
              subj = athlete_id) |> 
    select(y, argvals,subj)
 
#if (nrow(dati) == 0) next
  
dati_pred <- data.frame(y = rep(NA, nrow(dati) + k),
                          argvals = c(rep(NA, nrow(dati)), seq),
                          subj = rep(dati$subj[1], nrow(dati) + k)
  )
  
  dati_pred[1:nrow(dati), ] <- dati
  yhat <- predict(fit_face, dati_pred)
#Extract just the predictions on the grid  
  Ord <- (nrow(dati) + 1):(nrow(dati) + k)
  raw <- yhat$y.pred[Ord]

# Check if prediction is valid
if (all(is.na(raw)) || any(raw[-1] == 0, na.rm = TRUE)) next

# Compute percent change: (current - previous)/previous * 100
pc <- c(NA, diff(raw) / head(raw, -1)) * 100

# Save to matrix, dropping the initial NA
PercentChange_mat[i, ] <- pc[-1]
}

# conduct clustering of predicted functions from fpca sparse(k-means clustering)
set.seed(20240805)
cl_kmeans <- kmeans(PercentChange_mat, centers = 3, nstart = 25)
cl_ind <- cl_kmeans$cluster
cl_cen <- cl_kmeans$centers

# plot the predicted aging curves together with the estimated clustering as well as centers of the clusters
par(mar = c(4, 4, 1, 1))
colset <- c("#E69A8DFF", "#F6D55C", "#2A9D8F", "#5F4B8BFF")
plot(NULL, xlim = range(seq), ylim = c(-20, 20),
     xlab = "Performance Age (years)", ylab = "Percent Change in IAAF Score", bty = "n")

for(i in 1:n){
  lines(seq[-1], PercentChange_mat[i,], col = colset[cl_ind[i]])
}

for(j in 1:3){
  lines(seq[-1], cl_cen[j,], col = "black", lwd = 3)
}
legend("topright", legend = paste("Cluster", 1:3),
       col = colset, lty = 1, lwd = 2, bty = "n")

```

## Cluster percent-change by event type

```{r}
# CLuster by event type based on iaaf-point percent change 
# Filter and structure the dataset
fd_data_by_event <- athlete_bests_points |> 
  filter(!is.na(points), !is.na(age_years)) |> 
  mutate(
    athlete_id = paste(athlete_link, event),
    performance_age = age_years,
    iaaf_score = points,
    event_type = case_match(
       event,
      c("100m", "200m", "400 Metres", "400m", "400mH", "100mH", "110mH") ~ "Sprints",
      c("800m", "1500m", "3000m SC") ~ "Middle Distance",
      c("5000m", "10000m") ~ "Long Distance",
      c("Hept.", "Dec.") ~ "Combined Events",
      c("HJ", "LJ", "TJ", "PV") ~ "Jumps",
      c("SP", "DT", "HT", "JT") ~ "Throws",
      c("10,000mW", "20,000mW", "50,000mW", "Road Marathon") ~ "Road Races")
  ) |> 
  select(athlete_id, performance_age, iaaf_score, event_type)
#choose events
event_types <- unique(fd_data_by_event$event_type)
cluster_results <- list()
```
